\pkg{fitdistrplus}: An \proglang{R} Package for Fitting Distributions

### 2013-05-10 JSS/fitdistrplus/paper2JSS.Rnw
few changes to respect JSS requirement (to continue and check)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/riskassessment@248 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

Few packages on \sigle{CRAN} provide estimation procedures for any user-supplied parametric distribution 
and support different types of data. 
The \pkg{distrMod} package \citep{distrModJSS} provides an object-oriented (S4) 
implementation of probability models for non-censored data
and includes distribution fitting procedures for a given minimization criterion.
This package is based on the \proglang{R}-shipped function \code{mle} of the \pkg{stats4} package
producing S4 objects of class \code{"mle"} for which generic methods are 
implemented, e.g. \code{confint} and \code{logLik}.
In \pkg{fitdistrplus}, we chose to use the standard S3 class system for its understanding by most \proglang{R} users.
When designing the \pkg{fitdistrplus} package, we did not forget to implement
generic functions also available for S3 classes.
Finally, various other packages provide functions to estimate the mode, the moments or the L-moments of a distribution, see the reference manuals of  \pkg{modeest}, \pkg{lmomco} and \pkg{Lmoments} packages.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section[Fitting distributions]{Fitting distributions to continuous non-censored data}\label{fitnoncenscont}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Choice of candidate distributions}
\label{Choice}
For illustrating the use of various functions of the \pkg{fitdistrplus} package 
with continuous non-censored data, we will  first use
a data set named \code{groundbeef} which is included in our package.
This data set contains pointwise values of
serving sizes in grams, collected in a French survey, for ground beef patties consumed by 
children under 5 years old. It was used in a quantitative risk assessment 
published by \cite{Delignette08}.
%%% R code set default options for all R schunks
<<par4vignette, echo=FALSE, results=hide>>=
options(digits=4, SweaveHooks=list(fig=function() par(mar=c(5.1, 4.1, 1.1, 2.1))))
set.seed(1234)
@
%%% R code
<<datgroundbeef, echo=TRUE>>=
library(fitdistrplus)
data(groundbeef)
str(groundbeef)
@

@@

First of all, it is common to start with plots of the empirical distribution 
function and the histogram, which can be obtained with the \code{plotdist} function of the \pkg{fitdistrplus} package.
This function provides two plots (see Figure 1): the left-hand plot is the histogram (on a density scale) 
and the right-hand  plot the empirical cumulative distribution function (CDF).
%%% R code
<<figgroundbeef.echo, echo=TRUE, fig=FALSE, eval=FALSE>>=
plotdist(groundbeef$serving)
@

@@

\begin{figure}[htb!]
  \centering
  %%% R code
<<figgroundbeefplot, echo=FALSE, fig=TRUE, width=10, height=5, eps=FALSE>>=
plotdist(groundbeef$serving)
@
  \caption{Histogram and CDF plots of an empirical distribution for a continuous variable 
  (serving size from the \code{groundbeef} data set) as provided by the \code{plotdist} function}
  \label{plotdistcont}
\end{figure}

@@

\begin{figure}[htb]
  \centering
 %%% R code
<<descgroundbeefplot, echo=FALSE, fig=TRUE, width=6, height=6, eps=FALSE>>=
descdist(groundbeef$serving, boot=1000)
@
  \caption{Skewness-kurtosis plot for a continuous variable 
  (serving size from the \code{groundbeef} data set) as provided by the \code{descdist} function}
  \label{Cullenplotcont}
\end{figure}

@@

The \code{fitdist} function returns an S3 object of class \code{"fitdist"} 
for which \code{print}, \code{summary} and \code{plot} functions are provided.
The fit of a distribution using \code{fitdist} assumes that the corresponding 
\code{d}, \code{p}, \code{q} functions (standing respectively for the density, 
the distribution and the quantile functions) are defined.
Classical distributions are already defined in that way in the \pkg{stats} package,
e.g. \code{dnorm}, \code{pnorm} and \code{qnorm} for the normal distribution
(see \code{?Distributions}). Others
may be found in various packages (see the CRAN task view: Probability Distributions at \url{http://cran.r-project.org/web/views/Distributions.html}). 
Distributions not found in any package must be implemented by the user as
\code{d}, \code{p}, \code{q} functions.
In the call to \code{fitdist}, a distribution  has to be specified via
the argument \code{dist} either by the character string corresponding to its common root name
used in the names of \code{d}, \code{p}, \code{q} functions
(e.g. \code{"norm"} for the normal distribution) 
or by the density function itself, from which
the root name is extracted (e.g. \code{dnorm} for the normal distribution).
Numerical results returned by the \code{fitdist} function are 
(1) the parameter estimates, (2) the estimated standard errors (computed from the estimate
of the Hessian matrix at the maximum likelihood solution), (3) the loglikelihood, 
(4) Akaike and Bayesian information criteria (the so-called AIC and BIC), and (5) the correlation matrix
between parameter estimates.
Below is a call to the \code{fitdist} function to fit a Weibull distribution 
to the serving size from the \code{groundbeef} data set.
%%% R code
<<fitgroundbeef.weibull, echo = TRUE, fig = FALSE>>=
fw <- fitdist(groundbeef$serving, "weibull")
summary(fw)
@

The plot of an object of class \code{"fitdist"} provides 
four classical goodness-of-fit plots \citep{Cullen99} 
presented on Figure \ref{groundbeef:comp}: 
\begin{itemize}
\item a density plot representing the density function of the fitted distribution 
along with the histogram of the empirical distribution, 
\item a CDF plot of both the empirical distribution and the fitted distribution, 
\item a Q-Q plot
representing the empirical quantiles (y-axis) against the theoretical quantiles (x-axis)
\item a P-P plot representing the empirical distribution function evaluated at each data point (y-axis)
against the fitted distribution function (x-axis).
\end{itemize}

@@

\begin{figure}[htb!]
  \centering
<<fitgroundbeef, echo=FALSE, fig=TRUE, width=7, height=7, eps=FALSE>>=
par(mfrow=c(2, 2))
denscomp(list(fw,fln,fg), legendtext=c("Weibull", "lognormal", "gamma"))
qqcomp(list(fw,fln,fg), legendtext=c("Weibull", "lognormal", "gamma"))
cdfcomp(list(fw,fln,fg), legendtext=c("Weibull", "lognormal", "gamma"))
ppcomp(list(fw,fln,fg), legendtext=c("Weibull", "lognormal", "gamma"))
@ 
  \caption{Four Goodness-of-fit plots for various distributions fitted to continuous data
  (Weibull, gamma and lognormal
  distributions fitted to serving sizes from the \code{groundbeef} data set)  
  as provided by functions \code{denscomp}, \code{qqcomp}, \code{cdfcomp} and \code{ppcomp}}
\label{groundbeef:comp}  
\end{figure}

@@

The data set named  \code{endosulfan} will now be used to illustrate other features of 
the \pkg{fitdistrplus} package. This data set contains
acute toxicity values for the organochlorine pesticide endosulfan
(geometric mean of LC50 ou EC50 values in $\mu g.L^{-1}$),
tested on Australian and non-Australian laboratory-species 
%(arthropods, fish or nonarthropod invertebrates)
\citep{Hose04}. 
<<datendosulfan, echo=TRUE, eval=FALSE>>=
data(endosulfan)
str(endosulfan)
@
<<datendosulfan2, echo=FALSE, eval=TRUE>>=
data(endosulfan)
str(endosulfan, width = 70, strict.width = "cut")
@

@@

%%% R code
<<fitendo.echo, echo=TRUE, fig=FALSE>>=
ATV <-endosulfan$ATV
fendo.ln <- fitdist(ATV, "lnorm")
library(actuar)
fendo.ll <- fitdist(ATV, "llogis", start=list(shape=1, scale=500))
fendo.P <- fitdist(ATV, "pareto", start=list(shape=1, scale=500))
fendo.B <- fitdist(ATV, "burr", start=list(shape1=0.3, shape2=1, rate=1))
cdfcomp(list(fendo.ln, fendo.ll, fendo.P, fendo.B), xlogscale = TRUE,
      ylogscale = TRUE, legendtext = 
          c("lognormal", "loglogistic", "Pareto", "Burr"))
@
% qqcomp(list(fendo.ln, fendo.ll, fendo.P, fendo.B), xlogscale=TRUE, ylogscale=TRUE,
%       legendtext = c("lognormal","loglogistic","Pareto","Burr"))

@@

  \caption{CDF plot to compare the fit of four distributions to
  acute toxicity values of various organisms for the organochlorine pesticide endosulfan
  (\code{endosulfan} data set) as provided by the \code{cdfcomp} function, with CDF values
  in a logscale to emphasize discrepancies on the left tail}
\label{endo:comp}  
\end{figure}

@@

The computation of different goodness-of-fit statistics is proposed in the \pkg{fitdistrplus} package
in order to further compare fitted distributions.
The purpose of goodness-of-fit statistics aims to measure the distance between the
fitted parametric distribution and the empirical distribution: e.g. 
the distance between the fitted cumulative distribution function $F$ 
and the empirical distribution function $F_{n}$.
When fitting continuous distributions, three  goodness-of-fit statistics are classicaly considered: 
Cramer-von Mises, Kolmogorov-Smirnov and Anderson-Darling statistics \citep{Stephens86}.
Naming $x_{i}$ the $n$ observations of a continuous variable $X$ arranged in an ascending order,
Table \ref{tabKSCvMAD} gives the definition and the empirical estimate of the three considered 
goodness-of-fit statistics.
They can be computed using the function \code{gofstat}
as defined by Stephens \citep{Stephens86}.

@@

<<fendo.gof.print, echo=TRUE, fig=FALSE>>=
gofstat(list(fendo.ln, fendo.ll, fendo.P, fendo.B))
@

@@

%\newpage
As giving  more weight to distribution tails, 
the Anderson-Darling statistic is of special interest when it matters to equally 
emphasize the tails as well as the main body of a distribution.
This is often the case in risk assessment \citep{Cullen99,Vose10}. For this reason, 
this statistics is often used to select the best distribution among those fitted.
Nevertheless, this statistics should be used cautiously when comparing fits of various distributions. 
Keeping in mind that the weighting of each CDF quadratic difference 
depends on the parametric distribution in its definition (see Table \ref{tabKSCvMAD}), 
Anderson-Darling statistics computed for several distributions fitted on a same data set
are theoretically difficult to compare. Moreover, such a statistic, as Cramer-von Mises and
Kolmogorov-Smirnov ones, does not take into account the complexity of 
the model (i.e. parameter number). It is not
a problem when compared distributions are characterized by the same number of parameters, but
it could systematically promote the selection of the more complex distributions in the other case. 
Looking at classical penalized criteria based on the loglikehood (AIC, BIC) seems thus also interesting, especially
to discourage overfitting.

@@

\setkeys{Gin}{width=0.5\textwidth} 
\begin{figure}[htb!]
  \centering
<<fitBurrbootplot, echo=FALSE, fig=TRUE, width=6, height=6, eps=FALSE>>=
plot(bendo.B)
@
  \caption{Bootstrappped values of parameters for a fit of the Burr distribution
characterized by three parameters (example on the \code{endosulfan} data set) as provided 
by the plot of an object of class \code{"bootdist"}}
  \label{fig:bootstrap}
\end{figure}

@@

The bootstrap method can also be used to calculate confidence intervals on quantiles 
of the fitted distribution.
For this purpose, a generic \code{quantile} function is provided for class \code{bootdist}. 
By default, $95\%$ percentiles bootstrap confidence intervals of quantiles are provided.
Going back to the previous example from ecotoxicolgy, this function can be used to 
estimate the uncertainty associated to the HC5 estimation, for example from the previously fitted
Burr distribution to the \code{endosulfan} data set.
 %%% R code
<<fitATV.lnorm.quantile, echo=TRUE>>=
quantile(bendo.B, probs = 0.05)
@

@@

\begin{figure}[htb!]
  \centering
  %%% R code
<<mgegofcompplot, echo=FALSE, fig=TRUE, width=4, height=4, eps=FALSE>>=
cdfcomp(list(fendo.ln, fendo.ln.ADL, fendo.ln.AD2L),
xlogscale = TRUE, ylogscale = TRUE,
main = "Fitting a lognormal distribution",
legendtext = c("MLE","Left-tail AD", "Left-tail AD 2nd order"),
xlegend = "bottomright")
@
  \caption{Comparison of a lognormal distribution fitted by MLE
  and by MGE using  two different goodness-of-fit distances : left-tail Anderson-Darling and left-tail Anderson Darling of second order (example with the \code{endosulfan} data set)
  as provided by the \code{cdfcomp} function, with CDF values
  in a logscale to emphasize discrepancies on the left tail}
  \label{plotfitMGE}
\end{figure}

@@

%\subsubsection{Moment matching estimation}
%\label{MME}
%%%%%%%%%%
The moment matching estimation (MME) is another method commonly used to fit parametric 
distributions \citep{Vose10}. 
MME consists in finding the value of the parameter $\theta$  that equalizes the first theoretical 
raw moments of the parametric distribution  to the corresponding empirical raw moments as in Equation (\ref{moments}):
  \begin{equation}
    \label{moments}
    E(X^{k}|\theta)=\frac{1}{n}\sum_{i=1}^{n}x_{i}^{k} ,
  \end{equation}
for $k=1,\ldots,d$, with $d$ the number of parameters to estimate and
 $x_{i}$ the $n$ observations of variable $X$.
For moments of order greater than or equal to 2, it may  also be relevant to match centered moments.
Therefore, we match the moments given in Equation (\ref{centmoments}):
  \begin{equation}
    \label{centmoments}
    E(X\vert \theta) = \overline{x} ~,~
    E\left((X-E(X))^{k}|\theta\right)=m_k, \text{ for }  k=2,\ldots,d,
  \end{equation}
where $m_k$ denotes the empirical centered moments.
This method can be performed by setting the argument \code{method} to \code{"mme"} in the call to \code{fitdist}.
The estimate is computed by a closed-form formula for the following distributions:
normal, lognormal, exponential, Poisson, gamma, logistic, negative binomial,
geometric, beta and uniform distributions.
In this case, for distributions characterized by one parameter (geometric, Poisson and exponential),
this parameter is simply estimated by matching theoretical and observed means, and for distributions 
characterized by two parameters, these parameters are estimated by matching theoretical and observed 
means and variances \citep{Vose10}.
For other distributions, the equation of moments is solved numerically using 
the \code{optim} function  by minimizing the
sum of squared differences between observed and theoretical moments (see the \pkg{fitdistrplus} reference
manual
for technical details \citep{fitdistrplus}).

@@

The lognormal distribution is 
fitted to \code{danishuni} data set by matching moments implemented as a closed-form formula.
On the left-hand graph of Figure \ref{fig:danish:mme}, the fitted distribution functions obtained 
using the moment matching estimation (MME) and  maximum likelihood estimation (MLE) methods are compared.
The MME method provides a more cautious estimation of the insurance risk 
as the MME-fitted distribution function (resp. MLE-fitted) underestimates (overestimates) 
the empirical distribution function for large values of claim amounts.

%%% R code
<<danish.mme, echo=TRUE, fig=FALSE>>=
data(danishuni)
str(danishuni)
fdanish.ln.MLE <- fitdist(danishuni$Loss, "lnorm")
fdanish.ln.MME <- fitdist(danishuni$Loss, "lnorm", method="mme", order=1:2)
cdfcomp(list(fdanish.ln.MLE, fdanish.ln.MME), 
        legend=c("lognormal MLE", "lognormal MME"), 
        main="Fitting a lognormal distribution",
        xlogscale=TRUE, datapch=20)
@

\setkeys{Gin}{width=0.9\textwidth} %default
\begin{figure}[htb!]
  \centering
  %%% R code
<<danishmmeplot, echo=FALSE, fig=TRUE, width=7, height=3.5, eps=FALSE>>=
library(actuar)
fdanish.P.MLE <- fitdist(danishuni$Loss, "pareto", 
start=c(shape=10, scale=10), lower=2+1e-6, upper=Inf)
memp <- function(x, order) sum(x^order)/length(x)
fdanish.P.MME <- fitdist(danishuni$Loss, "pareto", method="mme", order=1:2, 
memp="memp", start=c(shape=10, scale=10), lower=c(2+1e-6,2+1e-6), 
upper=c(Inf,Inf))
par(mfrow=c(1, 2))
cdfcomp(list(fdanish.ln.MLE, fdanish.ln.MME), 
        legend=c("lognormal MLE", "lognormal MME"), main="Fitting a lognormal distribution",
        xlogscale=TRUE, datapch=20)
cdfcomp(list(fdanish.P.MLE, fdanish.P.MME), 
        legend=c("Pareto MLE", "Pareto MME"), main="Fitting a Pareto distribution",
        xlogscale=TRUE, datapch=20)
@
\caption{Comparison between MME and MLE when fitting a lognormal or a Pareto distribution 
to loss data from the \code{danishuni} data set}
\label{fig:danish:mme}
\end{figure}

@@

We use the implementation of the \pkg{actuar} package providing raw and centered moments
for that distribution (in addition to \code{d}, \code{p}, \code{q} and \code{r} functions \citep{actuar12}.
Fitting a heavy-tailed distribution for which the first and the second moments do not exist for
certain values of the shape parameter requires some cautiousness.
This is carried out by providing, for the optimization process, 
a lower and an upper bound for each parameter.
The code below calls the L-BFGS-B optimization method in \code{optim}, since this quasi-Newton
allows box constraints\footnote{That is what the B stands for.}.
We choose match moments defined in Equation (\ref{moments}), and so 
a function for computing the empirical raw moment (called \code{memp} in our example) is passed to \code{fitdist}.
For two-parameter distributions (i.e. $d=2$),  
Equations (\ref{moments}) and (\ref{centmoments}) are equivalent.

>>>>

Few packages on \sigle{CRAN} provide estimation procedures for any user-supplied parametric distribution 
and support different types of data. 
The \pkg{distrMod} package \citep{distrModJSS} provides an object-oriented (S4) 
implementation of probability models for non-censored data
and includes distribution fitting procedures for a given minimization criterion.
This package is based on the \proglang{R}-shipped function \code{mle} of the \pkg{stats4} package
producing S4 objects of class \code{"mle"} for which generic methods are 
implemented, e.g., \code{confint} and \code{logLik}.
In \pkg{fitdistrplus}, we chose to use the standard S3 class system for its understanding by most \proglang{R} users.
When designing the \pkg{fitdistrplus} package, we did not forget to implement
generic functions also available for S3 classes.
Finally, various other packages provide functions to estimate the mode, the moments or the L-moments of a distribution, see the reference manuals of  \pkg{modeest}, \pkg{lmomco} and \pkg{Lmoments} packages.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section[Fitting distributions]{Fitting distributions to continuous non-censored data}\label{fitnoncenscont}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Choice of candidate distributions}
\label{Choice}
For illustrating the use of various functions of the \pkg{fitdistrplus} package 
with continuous non-censored data, we will  first use
a data set named \code{groundbeef} which is included in our package.
This data set contains pointwise values of
serving sizes in grams, collected in a French survey, for ground beef patties consumed by 
children under 5 years old. It was used in a quantitative risk assessment 
published by \cite{Delignette08}.
%%% R code set default options for all R schunks
<<par4vignette, echo=FALSE, results=hide>>=
options(digits=4, prompt="R>", 
        SweaveHooks=list(fig=function() par(mar=c(5.1, 4.1, 1.1, 2.1))))
set.seed(1234)
@
%%% R code
<<datgroundbeef, echo=TRUE>>=
library("fitdistrplus")
data("groundbeef")
str(groundbeef)
@

@@

First of all, it is common to start with plots of the empirical distribution 
function and the histogram, which can be obtained with the \code{plotdist} function of the \pkg{fitdistrplus} package.
This function provides two plots (see Figure~\ref{plotdistcont}): the left-hand plot is the histogram (on a density scale) 
and the right-hand  plot the empirical cumulative distribution function (CDF).
%%% R code
<<figgroundbeef.echo, echo=TRUE, fig=FALSE, eval=FALSE>>=
plotdist(groundbeef$serving)
@

@@

\begin{figure}[htb!]
  \centering
  %%% R code
<<figgroundbeefplot, echo=FALSE, fig=TRUE, width=10, height=5, eps=FALSE>>=
plotdist(groundbeef$serving)
@
  \caption{Histogram and CDF plots of an empirical distribution for a continuous variable 
  (serving size from the \code{groundbeef} data set) as provided by the \code{plotdist} function.}
  \label{plotdistcont}
\end{figure}

@@

\begin{figure}[htb]
  \centering
 %%% R code
<<descgroundbeefplot, echo=FALSE, fig=TRUE, width=6, height=6, eps=FALSE>>=
descdist(groundbeef$serving, boot=1000)
@
  \caption{Skewness-kurtosis plot for a continuous variable 
  (serving size from the \code{groundbeef} data set) as provided by the \code{descdist} function.}
  \label{Cullenplotcont}
\end{figure}

@@

The \code{fitdist} function returns an S3 object of class \code{"fitdist"} 
for which \code{print}, \code{summary} and \code{plot} functions are provided.
The fit of a distribution using \code{fitdist} assumes that the corresponding 
\code{d}, \code{p}, \code{q} functions (standing respectively for the density, 
the distribution and the quantile functions) are defined.
Classical distributions are already defined in that way in the \pkg{stats} package,
e.g., \code{dnorm}, \code{pnorm} and \code{qnorm} for the normal distribution
(see \code{?Distributions}). Others
may be found in various packages (see the CRAN task view: Probability Distributions at \url{http://cran.r-project.org/web/views/Distributions.html}). 
Distributions not found in any package must be implemented by the user as
\code{d}, \code{p}, \code{q} functions.
In the call to \code{fitdist}, a distribution  has to be specified via
the argument \code{dist} either by the character string corresponding to its common root name
used in the names of \code{d}, \code{p}, \code{q} functions
(e.g., \code{"norm"} for the normal distribution) 
or by the density function itself, from which
the root name is extracted (e.g., \code{dnorm} for the normal distribution).
Numerical results returned by the \code{fitdist} function are 
(1) the parameter estimates, (2) the estimated standard errors (computed from the estimate
of the Hessian matrix at the maximum likelihood solution), (3) the loglikelihood, 
(4) Akaike and Bayesian information criteria (the so-called AIC and BIC), and (5) the correlation matrix
between parameter estimates.
Below is a call to the \code{fitdist} function to fit a Weibull distribution 
to the serving size from the \code{groundbeef} data set.
%%% R code
<<fitgroundbeef.weibull, echo = TRUE, fig = FALSE>>=
fw <- fitdist(groundbeef$serving, "weibull")
summary(fw)
@

The plot of an object of class \code{"fitdist"} provides 
four classical goodness-of-fit plots \citep{Cullen99} 
presented on Figure~\ref{groundbeef:comp}: 
\begin{itemize}
\item a density plot representing the density function of the fitted distribution 
along with the histogram of the empirical distribution, 
\item a CDF plot of both the empirical distribution and the fitted distribution, 
\item a Q-Q plot
representing the empirical quantiles (y-axis) against the theoretical quantiles (x-axis)
\item a P-P plot representing the empirical distribution function evaluated at each data point (y-axis)
against the fitted distribution function (x-axis).
\end{itemize}

@@

\begin{figure}[htb!]
  \centering
<<fitgroundbeef, echo=FALSE, fig=TRUE, width=7, height=7, eps=FALSE>>=
par(mfrow=c(2, 2))
denscomp(list(fw,fln,fg), legendtext=c("Weibull", "lognormal", "gamma"))
qqcomp(list(fw,fln,fg), legendtext=c("Weibull", "lognormal", "gamma"))
cdfcomp(list(fw,fln,fg), legendtext=c("Weibull", "lognormal", "gamma"))
ppcomp(list(fw,fln,fg), legendtext=c("Weibull", "lognormal", "gamma"))
@ 
  \caption{Four Goodness-of-fit plots for various distributions fitted to continuous data
  (Weibull, gamma and lognormal
  distributions fitted to serving sizes from the \code{groundbeef} data set)  
  as provided by functions \code{denscomp}, \code{qqcomp}, \code{cdfcomp} and \code{ppcomp}.}
\label{groundbeef:comp}  
\end{figure}

@@

The data set named  \code{endosulfan} will now be used to illustrate other features of 
the \pkg{fitdistrplus} package. This data set contains
acute toxicity values for the organochlorine pesticide endosulfan
(geometric mean of LC50 ou EC50 values in $\mu g.L^{-1}$),
tested on Australian and non-Australian laboratory-species 
%(arthropods, fish or nonarthropod invertebrates)
\citep{Hose04}. 
<<datendosulfan, echo=TRUE, eval=FALSE>>=
data("endosulfan")
str(endosulfan)
@
<<datendosulfan2, echo=FALSE, eval=TRUE>>=
data("endosulfan")
str(endosulfan, width = 70, strict.width = "cut")
@

@@

%%% R code
<<fitendo.echo, echo=TRUE, fig=FALSE>>=
ATV <-endosulfan$ATV
fendo.ln <- fitdist(ATV, "lnorm")
library("actuar")
fendo.ll <- fitdist(ATV, "llogis", start=list(shape=1, scale=500))
fendo.P <- fitdist(ATV, "pareto", start=list(shape=1, scale=500))
fendo.B <- fitdist(ATV, "burr", start=list(shape1=0.3, shape2=1, rate=1))
cdfcomp(list(fendo.ln, fendo.ll, fendo.P, fendo.B), xlogscale = TRUE,
      ylogscale = TRUE, legendtext = 
          c("lognormal", "loglogistic", "Pareto", "Burr"))
@
% qqcomp(list(fendo.ln, fendo.ll, fendo.P, fendo.B), xlogscale=TRUE, ylogscale=TRUE,
%       legendtext = c("lognormal","loglogistic","Pareto","Burr"))

@@

  \caption{CDF plot to compare the fit of four distributions to
  acute toxicity values of various organisms for the organochlorine pesticide endosulfan
  (\code{endosulfan} data set) as provided by the \code{cdfcomp} function, with CDF values
  in a logscale to emphasize discrepancies on the left tail.}
\label{endo:comp}  
\end{figure}

@@

The computation of different goodness-of-fit statistics is proposed in the \pkg{fitdistrplus} package
in order to further compare fitted distributions.
The purpose of goodness-of-fit statistics aims to measure the distance between the
fitted parametric distribution and the empirical distribution: e.g., 
the distance between the fitted cumulative distribution function $F$ 
and the empirical distribution function $F_{n}$.
When fitting continuous distributions, three  goodness-of-fit statistics are classicaly considered: 
Cramer-von Mises, Kolmogorov-Smirnov and Anderson-Darling statistics \citep{Stephens86}.
Naming $x_{i}$ the $n$ observations of a continuous variable $X$ arranged in an ascending order,
Table \ref{tabKSCvMAD} gives the definition and the empirical estimate of the three considered 
goodness-of-fit statistics.
They can be computed using the function \code{gofstat}
as defined by Stephens \citep{Stephens86}.

@@

<<fendo.gof.print, echo=TRUE, fig=FALSE>>=
gofstat(list(fendo.ln, fendo.ll, fendo.P, fendo.B),
        fitnames = c("lnorm","llogis","Pareto","Burr"))
@

@@

%\newpage
As giving  more weight to distribution tails, 
the Anderson-Darling statistic is of special interest when it matters to equally 
emphasize the tails as well as the main body of a distribution.
This is often the case in risk assessment \citep{Cullen99,Vose10}. For this reason, 
this statistics is often used to select the best distribution among those fitted.
Nevertheless, this statistics should be used cautiously when comparing fits of various distributions. 
Keeping in mind that the weighting of each CDF quadratic difference 
depends on the parametric distribution in its definition (see Table \ref{tabKSCvMAD}), 
Anderson-Darling statistics computed for several distributions fitted on a same data set
are theoretically difficult to compare. Moreover, such a statistic, as Cramer-von Mises and
Kolmogorov-Smirnov ones, does not take into account the complexity of 
the model (i.e., parameter number). It is not
a problem when compared distributions are characterized by the same number of parameters, but
it could systematically promote the selection of the more complex distributions in the other case. 
Looking at classical penalized criteria based on the loglikehood (AIC, BIC) seems thus also interesting, especially
to discourage overfitting.

@@

\setkeys{Gin}{width=0.5\textwidth} 
\begin{figure}[htb!]
  \centering
<<fitBurrbootplot, echo=FALSE, fig=TRUE, width=6, height=6, eps=FALSE>>=
plot(bendo.B)
@
  \caption{Bootstrappped values of parameters for a fit of the Burr distribution
characterized by three parameters (example on the \code{endosulfan} data set) as provided 
by the plot of an object of class \code{"bootdist"}.}
  \label{fig:bootstrap}
\end{figure}

@@

The bootstrap method can also be used to calculate confidence intervals on quantiles 
of the fitted distribution.
For this purpose, a generic \code{quantile} function is provided for class \code{bootdist}. 
By default, $95\%$ percentiles bootstrap confidence intervals of quantiles are provided.
Going back to the previous example from ecotoxicolgy, this function can be used to 
estimate the uncertainty associated to the HC5 estimation, for example from the previously fitted
Burr distribution to the \code{endosulfan} data set.
 %%% R code
<<fitATV.lnorm.quantile, echo=TRUE,results=hide>>=
quantile(bendo.B, probs = 0.05)
@
\begin{footnotesize}
<<fitATV.lnorm.quantileb, echo=FALSE>>=
quantile(bendo.B, probs = 0.05)
@
\end{footnotesize}

@@

\begin{figure}[htb!]
  \centering
  %%% R code
<<mgegofcompplot, echo=FALSE, fig=TRUE, width=4, height=4, eps=FALSE>>=
cdfcomp(list(fendo.ln, fendo.ln.ADL, fendo.ln.AD2L),
xlogscale = TRUE, ylogscale = TRUE,
main = "Fitting a lognormal distribution",
legendtext = c("MLE","Left-tail AD", "Left-tail AD 2nd order"),
xlegend = "bottomright")
@
  \caption{Comparison of a lognormal distribution fitted by MLE
  and by MGE using  two different goodness-of-fit distances : left-tail Anderson-Darling and left-tail Anderson Darling of second order (example with the \code{endosulfan} data set)
  as provided by the \code{cdfcomp} function, with CDF values
  in a logscale to emphasize discrepancies on the left tail.}
  \label{plotfitMGE}
\end{figure}

@@

%\subsubsection{Moment matching estimation}
%\label{MME}
%%%%%%%%%%
The moment matching estimation (MME) is another method commonly used to fit parametric 
distributions \citep{Vose10}. 
MME consists in finding the value of the parameter $\theta$  that equalizes the first theoretical 
raw moments of the parametric distribution  to the corresponding empirical raw moments as in Equation~(\ref{moments}):
  \begin{equation}
    \label{moments}
    E(X^{k}|\theta)=\frac{1}{n}\sum_{i=1}^{n}x_{i}^{k} ,
  \end{equation}
for $k=1,\ldots,d$, with $d$ the number of parameters to estimate and
 $x_{i}$ the $n$ observations of variable $X$.
For moments of order greater than or equal to 2, it may  also be relevant to match centered moments.
Therefore, we match the moments given in Equation~(\ref{centmoments}):
  \begin{equation}
    \label{centmoments}
    E(X\vert \theta) = \overline{x} ~,~
    E\left((X-E(X))^{k}|\theta\right)=m_k, \text{ for }  k=2,\ldots,d,
  \end{equation}
where $m_k$ denotes the empirical centered moments.
This method can be performed by setting the argument \code{method} to \code{"mme"} in the call to \code{fitdist}.
The estimate is computed by a closed-form formula for the following distributions:
normal, lognormal, exponential, Poisson, gamma, logistic, negative binomial,
geometric, beta and uniform distributions.
In this case, for distributions characterized by one parameter (geometric, Poisson and exponential),
this parameter is simply estimated by matching theoretical and observed means, and for distributions 
characterized by two parameters, these parameters are estimated by matching theoretical and observed 
means and variances \citep{Vose10}.
For other distributions, the equation of moments is solved numerically using 
the \code{optim} function  by minimizing the
sum of squared differences between observed and theoretical moments (see the \pkg{fitdistrplus} reference
manual
for technical details \citep{fitdistrplus}).

@@

The lognormal distribution is 
fitted to \code{danishuni} data set by matching moments implemented as a closed-form formula.
On the left-hand graph of Figure~\ref{fig:danish:mme}, the fitted distribution functions obtained 
using the moment matching estimation (MME) and  maximum likelihood estimation (MLE) methods are compared.
The MME method provides a more cautious estimation of the insurance risk 
as the MME-fitted distribution function (resp. MLE-fitted) underestimates (overestimates) 
the empirical distribution function for large values of claim amounts.

%%% R code
<<danish.mme, echo=TRUE, fig=FALSE>>=
data("danishuni")
str(danishuni)
fdanish.ln.MLE <- fitdist(danishuni$Loss, "lnorm")
fdanish.ln.MME <- fitdist(danishuni$Loss, "lnorm", method="mme", order=1:2)
cdfcomp(list(fdanish.ln.MLE, fdanish.ln.MME), 
        legend=c("lognormal MLE", "lognormal MME"), 
        main="Fitting a lognormal distribution",
        xlogscale=TRUE, datapch=20)
@

\setkeys{Gin}{width=0.9\textwidth} %default
\begin{figure}[htb!]
  \centering
  %%% R code
<<danishmmeplot, echo=FALSE, fig=TRUE, width=7, height=3.5, eps=FALSE>>=
library("actuar")
fdanish.P.MLE <- fitdist(danishuni$Loss, "pareto", 
start=c(shape=10, scale=10), lower=2+1e-6, upper=Inf)
memp <- function(x, order) sum(x^order)/length(x)
fdanish.P.MME <- fitdist(danishuni$Loss, "pareto", method="mme", order=1:2, 
memp="memp", start=c(shape=10, scale=10), lower=c(2+1e-6,2+1e-6), 
upper=c(Inf,Inf))
par(mfrow=c(1, 2))
cdfcomp(list(fdanish.ln.MLE, fdanish.ln.MME), 
        legend=c("lognormal MLE", "lognormal MME"), main="Fitting a lognormal distribution",
        xlogscale=TRUE, datapch=20)
cdfcomp(list(fdanish.P.MLE, fdanish.P.MME), 
        legend=c("Pareto MLE", "Pareto MME"), main="Fitting a Pareto distribution",
        xlogscale=TRUE, datapch=20)
@
\caption{Comparison between MME and MLE when fitting a lognormal or a Pareto distribution 
to loss data from the \code{danishuni} data set.}
\label{fig:danish:mme}
\end{figure}

@@

We use the implementation of the \pkg{actuar} package providing raw and centered moments
for that distribution (in addition to \code{d}, \code{p}, \code{q} and \code{r} functions \citep{actuar12}.
Fitting a heavy-tailed distribution for which the first and the second moments do not exist for
certain values of the shape parameter requires some cautiousness.
This is carried out by providing, for the optimization process, 
a lower and an upper bound for each parameter.
The code below calls the L-BFGS-B optimization method in \code{optim}, since this quasi-Newton
allows box constraints\footnote{That is what the B stands for.}.
We choose match moments defined in Equation~(\ref{moments}), and so 
a function for computing the empirical raw moment (called \code{memp} in our example) is passed to \code{fitdist}.
For two-parameter distributions (i.e., $d=2$),  
Equations~(\ref{moments}) and (\ref{centmoments}) are equivalent.
